Abstractions for Multi-Sorted Substitutions

### 2024-03-18 paper/paper.tex
Update.

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    just works out of the box for both deriving and instantiating the
    \ACode{\ACon{TTraversal}}.
    Furthermore, the \ACode{\ACon{CKit}} abstraction turned out
    to be useful to prove confluence, allowing to unify certain
    lemmas, which otherwise need to be proved for both renamings and
    substitutions.
    % e.g. that for two terms \ACode{t} and \ACode{t'}
    % with \ACode{t \ACon{↪} t'} it follows that \ACode{(t \AField{⋯} ϕ)
    % \ACon{↪} (t' \AField{⋯} ϕ)} for any map \ACode{ϕ}.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution preserves typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we found an encoding, which still allows us to use the
    the \ACode{\ACon{TKit}} abstraction.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With that encoding substitution preserves typing is generally true
    again, enabling the use of the framework's typing contexts and
    \ACode{\ACon{TKit}} machinery.
    When replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, then the resulting term is still
    well-typed, but in a context with an unsatisfiable constraint.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}

@@

  As the amount of related work is rather large, we focus on work that
  is closely related to ours and refer to other papers for the
  broader
  picture\cite{DBLP:conf/cpp/StarkSK19,DBLP:journals/pacmpl/AllaisA0MM18}. In
  particular, we discuss abstractions for renamings
  and substitutions, comparable extrinsically typed approaches, but also
  intrinsically typed approaches that could be used as a basis for
  multi-sorted syntax, which as mentioned before is intrinsically typed
  in disguise.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases kits are formulated for intrinsic typing
  and scenarios with polymorphism are not yet considered.

@@

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  Compared to our work, the syntax they generate takes the form of
  what we described as \emph{unsorted syntax} in
  Section~\ref{sec:syntax}, i.e.\ different syntactic categories are
  described by different types with different amount of indices for
  variable counts. To eliminate the need for interaction lemmas, they
  define the notion of \emph{vectorized substitution}, which combines
  the individual substitutions by putting them in a vector.

  Both works do not provide machinery for type preservation and do not
  unify renaming and substitution (although the latter is listed as
  future work). We believe their amazing work of creating a decision
  procedure for vectorized substitutions should also translate to our
  setting with multi-sorted substitutions.

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instantiate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics, how the four composition lemmas follow from
  their generic fusion lemma, and also provide an abstraction to unify
  renamings and substitutions.
  They show how to use their framework for both intrinsic and
  extrinsic typing, but have not yet explored the realm of polymorphism.

  This work is orthogonal, but compatible to our approach.
  We can instantiate their system using the hidden intrinsic type
  system of our multi-sorted syntax, but unfortunately then their
  semantics abstraction would only work with terms, which are untyped
  with respect to the actual extrinsic type system.
  We think it is worthwile to explore if their semantics abstraction
  can also be generalized to multi-sorted typing relations.

@@

  \subsection{Pure Type Systems}
  Pure Type Systems\cite{DBLP:conf/esop/BarendregtH90, berarditowards, terlouw1989een}
  describe a class of typed lambda calculi parameterized over a set of
  sorts, dependencies between sorts, and rules for quantification.

  While pure type systems may seem very similar to multi-sorted
  syntax, they are actually quite different:
  \begin{itemize}
  \item 
    In multi-sorted syntax, sorts describe syntactic categories of terms. Terms
    of different sorts are kept syntactically different, e.g. the set of expressions
    \ACode{S ⊢ 𝕖} and types \ACode{S ⊢ 𝕥} in our System F example.
  \item 
    In pure type systems, sorts are universe types, e.g.\ like
    \ACode{Set} and \ACode{Set₁} in Agda.
    Terms which have different sorts as types, do still belong to the
    same syntactic category.
  \end{itemize}

  We can use our framework to model pure type systems, but the sorts of pure type systems
  do not correspond to the sorts of the multi-sorted syntax. The multi-sorted syntax
  has just a single sort that describes all terms, as it is common for
  dependently typed languages.

  \section{Conclusion}
  \label{sec:conclusion}
  We have presented a framework for deriving untyped substitution and its properties, and
  and abstraction for proving type preservation of substitution with a single lemma.

  Using extrinsic typing allowed us to unify the renamings and
  substitutions for variables of arbitrary many syntactic categories.

  Furthermore, extrinsic typing allowed us treat untyped substitution
  and its properties separately from typing and type preservation.
  This has the benefit that the former can be derived automatically for virtually
  any object language, while the latter is more domain specific and
  can be handled independently.

  We believe our framework is a viable alternative to intrinsic
  typing, as the duplication is automated away.

  We believe our framework is particularly useful for polymorphic languages,
  where intrinsic typing requires necessarily to deal with multiple substitutions
  and their interactions. Furthermore, in more complex scenarios, intrinsic typing
  can easily lead to subst-hell, which is avoided with extrinsic typing.

  We belive the core limitation of our framework are the typing kits, which
  require a classical ternary typing relation. However, this problem is hard
  to avoid as typing relations can become arbitrarily complicated.
  To adapt the framework to more complicated type systems, only the typing
  kits and typing traversals need to be extended, which we have successfully
  done for linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.

@@

  \clearpage
  \appendix
  \onecolumn

  \section{Omitted Proofs}
  \label{sec:proofs}

  \subsection{Proofs for Section~\ref{sec:maps}}
  \LibCode\KIdLift
  \LibCode\KIdLiftProof

  % \AppCode\FTraversalId
  % \AppCode\FTraversalIdProofInteresting
  % \LibCode\FTraversalIdProofRest

  \subsection{Proofs for Section~\ref{sec:composition}}
  \LibCode\KCKitAp
  \LibCode\KCKitApProof

  \LibCode\KDistLiftCompose
  \LibCode\KDistLiftComposeProof

  \LibCode\KCommLiftWeaken
  \LibCode\KCommLiftWeakenProof

  \LibCode\KCommLiftWeakenTraverse
  \LibCode\KCommLiftWeakenTraverseProof

  \LibCode\KWeakenCancelsSingle
  \LibCode\KWeakenCancelsSingleProof

  \LibCode\KWeakenCancelsSingleTraverse
  \LibCode\KWeakenCancelsSingleTraverseProof

  \LibCode\KDistLiftSingle
  \LibCode\KDistLiftSingleProof

  \LibCode\KDistLiftSingleTraverse
  \LibCode\KDistLiftSingleTraverseProof

  % \AppCode\FFusion
  % \AppCode\FFusionProofInteresting
  % \AppCode\FFusionProofRest

  \subsection{Proofs for Section~\ref{sec:typing}}
  \LibCode\KLiftTyping
  \LibCode\KLiftTypingProof

  \LibCode\KSingleTyping
  \LibCode\KSingleTypingProof

  % \AppCode\FSubjectReduction
  % \AppCode\FSubjectReductionProofInteresting

  % \section{Generic Class of Object Languages}
  % \label{sec:generic-class}

  % \section{System F with Reflection}
  % \label{sec:reflection-example}

  % \section{Map Equality Without Functional Extensionality}
  % \label{sec:fun-ext}

  % \section{Features of the Actual Implementation}
  % \label{sec:features}

>>>>

  \section{Case Studies}
  \label{sec:casestudies}
  Using our full implementation of the framework, we proved subject
  reduction for the following object languages:
  \begin{itemize}
  \item
    For lambda calculus with dependent function types, the framework
    works out of the box for both deriving untyped substitution
    and instantiating the \ACode{\ACon{TTraversal}}.
    In the proof of confluence, the \ACode{\ACon{CKit}} abstraction
    allowed us to unify lemmas about the reduction of renamings and
    substitutions.
  \item
    For System F with subtyping the main challenges are how to represent
    subtyping constraints and how to deal with the fact that
    substitution-preserves-typing is not generally true, as type variables
    have subtyping bounds that need to be respected.
    While it would be possible to use our framework only to derive
    untyped substitution and define typing contexts and type preservation
    lemmas by hand, we instead used an encoding that allows us to use the
    the \ACode{\ACon{TKit}} abstraction directly.
    Instead of binding a type variable with a subtyping constraint
    \ACode{α <: t}, we first bind the type variable as \ACode{α : ★}, and
    then bind the constraint as \ACode{c ∶ (α <: t)}. This description is
    similar to first-class constraints, but restricted enough to be
    isomorphic to the original formalization, as constraint variables
    cannot be accesses by the user.
    With this encoding, substitution-preserves-typing is generally true
    again: replacing a type variable \ACode{α <: t} with a type \ACode{t'},
    which is not a subtype of \ACode{t}, results in a term that is still
    well-typed, but in a context with an unsatisfiable constraint \ACode{t' <: t}.
  \item
    Object languages with pattern matching can be encoded by adding
    the sorts of the variables bound by a pattern to the pattern sort
    \ACode{\ACon{𝕡}} itself:
    \begin{quote}
    \ACode{\ACon{𝕡} \AKw{:} \ACon{List (Sort Var)} \AKw{→} \ACon{Sort NoVar}} 
    \end{quote}
    This allows the clause of a \ACode{\ACon{match}} expression to
    read the bound variables from the pattern sort and make them
    available as free variables in its body expression.
  \end{itemize}

@@

  As the amount of related work is rather large, we focus on work that
  is closely related to ours and refer to other papers for the
  broader
  picture\cite{DBLP:conf/cpp/StarkSK19,DBLP:journals/pacmpl/AllaisA0MM18}.
  % In
  % particular, we discuss abstractions for renamings
  % and substitutions, comparable extrinsically typed approaches, but also
  % intrinsically typed approaches that could be used as a basis for
  % multi-sorted syntax, which as mentioned before is intrinsically typed
  % in disguise.

  \subsection{Variable Binding}
  There is a plethora of different methods for representing
  variable binders in the literature:
  de Bruijn\cite{scholar:de1972lambda},
  co de Bruijn\cite{DBLP:journals/corr/abs-1807-04085},
  locally nameless\cite{DBLP:journals/jar/Chargueraud12},
  locally named\cite{DBLP:conf/tlca/McKinnaP93},
  higher order abstract syntax\cite{DBLP:conf/pldi/PfenningE88} and
  its parametric variant\cite{DBLP:conf/icfp/Chlipala08},
  nominal logic\cite{DBLP:journals/tcs/UrbanPG04},
  shifted names\cite{scholar:dolansyntax},
  nameless painless\cite{DBLP:conf/icfp/Pouillard11}, and
  scope graphs\cite{DBLP:journals/pacmpl/AntwerpenPRV18}.
  Many of these representations have been studied in solutions to the POPLMark
  challenges\cite{DBLP:conf/tphol/AydemirBFFPSVWWZ05, DBLP:journals/jfp/AbelAHPMSS19}.

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases kits are formulated for intrinsic typing
  and scenarios with polymorphism are not considered.

@@

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  Compared to our work, the syntax they generate takes the form of
  what we described as \emph{unsorted syntax} in
  Section~\ref{sec:syntax}, i.e.\ different syntactic categories are
  described by different types with different amount of indices for
  variable counts. To eliminate the need for interaction lemmas, they
  define the notion of \emph{vectorized substitution}, which combines
  the individual substitutions by putting them in a vector.
  We believe their great work of creating a decision procedure for
  vectorized substitutions should also translate to our setting with
  multi-sorted substitutions.

  \emph{Needle and Knot}\cite{DBLP:conf/esop/KeuchelWS16} is a code
  generator for unscoped syntax with de Bruijn indices. They generate
  substitution and interaction lemmas for single-pointed substitution
  for languages with multiple variable sorts and binders that bind
  lists of variables.

  All of the above work does not provide machinery to model typing
  and type preservation and does not unify renaming and substitution and
  their compositions.
  Hence, type preservation needs to be modeled manually and
  individually for renamings and substitutions.

@@

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instantiate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics, how the four composition lemmas follow from
  their generic fusion lemma, and also provide an abstraction to unify
  renamings and substitutions.
  They show how to use their framework for both intrinsic and
  extrinsic typing, but are missing a story for polymorphism.

  With only a slight modification to their framework, we can instantiate it
  for multi-sorted syntax, enabling the definition of polymorphic languages.
  However, as the intrinsic typing is then
  used to ensure syntactic sorts, the semantic abstractions talk about untyped
  terms, so typing and type preservation lemmas have to be modeled entirely manually.
  We believe it would be worthwhile to explore how their semantics
  abstractions can be lifted to typing relations similar to how our typing kits lift
  regular kits from terms to typing relations.

@@

  \subsection{Pure Type Systems}
  Pure Type Systems\cite{DBLP:conf/esop/BarendregtH90, berarditowards, terlouw1989een}
  describe a class of typed lambda calculi parameterized over a set of
  sorts, dependencies between sorts, and rules for quantification.
  While pure type systems may seem very similar to multi-sorted
  syntax, they are actually quite different:
  \begin{itemize}
  \item 
    In multi-sorted syntax, sorts describe syntactic categories of terms. Terms
    of different sorts are kept syntactically different, e.g. the set of expressions
    \ACode{S ⊢ 𝕖} and types \ACode{S ⊢ 𝕥} in our System F example.
  \item 
    In pure type systems, sorts are universe types, e.g.\ like
    \ACode{Set} in Agda or \ACode{Prop} in Coq.
    Terms which have different sorts as types, do still belong to the
    same syntactic category.
  \end{itemize}

  We can model pure type systems as a multi-sorted syntax with a single sort,
  where the sorts of the pure type system occur as terms representing universe types.

  % We can use our framework to model pure type systems, but the sorts of pure type systems
  % do not correspond to the sorts of the multi-sorted syntax. The multi-sorted syntax
  % has just a single sort that describes all terms, as it is common for
  % dependently typed languages.

  \section{Conclusion}
  \label{sec:conclusion}
  % We have presented a framework for deriving untyped substitution and its properties, and
  % and abstraction for proving type preservation of substitution with a single lemma.

  % Using extrinsic typing allowed us to unify the renamings and
  % substitutions for variables of arbitrary many syntactic categories.

  % Furthermore, extrinsic typing allowed us treat untyped substitution
  % and its properties separately from typing and type preservation.
  % This has the benefit that the former can be derived automatically for virtually
  % any object language, while the latter is more domain specific and
  % can be handled independently.

  % We believe our framework is a viable alternative to intrinsic
  % typing, as the duplication is automated away.

  % We believe our framework is particularly useful for polymorphic languages,
  % where intrinsic typing requires necessarily to deal with multiple substitutions
  % and their interactions. Furthermore, in more complex scenarios, intrinsic typing
  % can easily lead to subst-hell, which is avoided with extrinsic typing.

  % We belive the core limitation of our framework are the typing kits, which
  % require a classical ternary typing relation. However, this problem is hard
  % to avoid as typing relations can become arbitrarily complicated.
  % To adapt the framework to more complicated type systems, only the typing
  % kits and typing traversals need to be extended, which we have successfully
  % done for linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  % general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.

  % - Main use case: polymorphism / multiple syntactic categories / dependent types
  % - Limitations and Flexibility with only extending typing kits
  % - Comparison with intrinsic typing
  %   - Deriving untyped substitution is on par with intrinsic
  %   - Subst-Hell / Interaction lemmas unavoidable

  While intrinsic typing allows to unify definitions with their type
  preservation proofs, extrinsic typing allows to unify
  substitutions across different syntactic categories, as we have demonstrated.
  We believe this makes extrinsic typing more suitable for polymorphic languages:
  with extrinsic typing, the definitions and lemmas for untyped
  substitutions can be automated for virtually any object language,
  whereas with intrinsic typing, it seems impossible to unify
  substitutions across syntactic categories, because an intrinsically
  typed multi-sorted syntax would require to index terms with terms,
  which is the same problem that occurs with giving intrinsic types to a dependently typed language.

  Our framework is able to automatically derive definitions and lemmas
  about untyped substitution, and provides an abstraction for proving
  type preservation of substitution with a single lemma (\ACode{\AField{\_⊢⋯\_}}).

  Similarly to frameworks with intrinsic typing, the main limitation in our work is the shape of
  typing relations: our typing kits can only model classical ternary typing relations.
  To adapt our framework to more complicated typing relations, the machinery for untyped substitution
  can be reused, but the typing kits and typing traversals need to be modified.
  We found that this works surprisingly well in practice,
  where we have already made extended typing kits to support
  linear typing ala Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} and
  general substructural typing ala Licata et al\cite{DBLP:conf/rta/LicataSR17}.
  In both cases, the typing relation has a fourth component which models usage restrictions
  on the type context.

  % In our opinion, our framework makes a case for extrinsic typing for
  % polymorphic, languages as the

@@

  \clearpage
  % \appendix
  % \onecolumn

  % \section{Omitted Proofs}
  % \label{sec:proofs}

  % \subsection{Proofs for Section~\ref{sec:maps}}
  % \LibCode\KIdLift
  % \LibCode\KIdLiftProof

  % % \AppCode\FTraversalId
  % % \AppCode\FTraversalIdProofInteresting
  % % \LibCode\FTraversalIdProofRest

  % \subsection{Proofs for Section~\ref{sec:composition}}
  % \LibCode\KCKitAp
  % \LibCode\KCKitApProof

  % \LibCode\KDistLiftCompose
  % \LibCode\KDistLiftComposeProof

  % \LibCode\KCommLiftWeaken
  % \LibCode\KCommLiftWeakenProof

  % \LibCode\KCommLiftWeakenTraverse
  % \LibCode\KCommLiftWeakenTraverseProof

  % \LibCode\KWeakenCancelsSingle
  % \LibCode\KWeakenCancelsSingleProof

  % \LibCode\KWeakenCancelsSingleTraverse
  % \LibCode\KWeakenCancelsSingleTraverseProof

  % \LibCode\KDistLiftSingle
  % \LibCode\KDistLiftSingleProof

  % \LibCode\KDistLiftSingleTraverse
  % \LibCode\KDistLiftSingleTraverseProof

  % % \AppCode\FFusion
  % % \AppCode\FFusionProofInteresting
  % % \AppCode\FFusionProofRest

  % \subsection{Proofs for Section~\ref{sec:typing}}
  % \LibCode\KLiftTyping
  % \LibCode\KLiftTypingProof

  % \LibCode\KSingleTyping
  % \LibCode\KSingleTypingProof

  % % \AppCode\FSubjectReduction
  % % \AppCode\FSubjectReductionProofInteresting

  % % \section{Generic Class of Object Languages}
  % % \label{sec:generic-class}

  % % \section{System F with Reflection}
  % % \label{sec:reflection-example}

  % % \section{Map Equality Without Functional Extensionality}
  % % \label{sec:fun-ext}

  % % \section{Features of the Actual Implementation}
  % % \label{sec:features}
